Secure Composition \\ of Robust and Optimising Compilers

### 2024-05-27 csf2/main.tex
Space opts

Properties $\pi$ are sets of traces of admissible program behaviours, ascribing what said property considers valid.
The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}.
A class is simply a set of properties and for the class of safety properties, it is decidable whether a trace satisfies a safety property with just a finite trace prefix.
The compositionality framework (\Cref{sec:sequential}) presented in this paper considers arbitrary classes, while the case-study (\Cref{sec:casestud:rtp}) fixes them to concrete instances.
As an example, consider a trace describing an interaction with a memory where the deallocation of an address $\loc$ precedes a read at that address in memory: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$.
This program behaviour is insecure w.r.t. a canonical notion of (temporal) memory safety dictating no use-after-frees of pointers~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The previous finite trace prefix is enough to decide that the trace does not satisfy \gls*{tms} and there is no way to append events to this prefix which would result in the trace being admissible.
In the following, the execution of a whole program $\wholeprogvar$ that terminates in state $\runtimetermvar$ according to the language semantics and produces trace $\trace$ is written as $\progstepto{\wholeprogvar}{\runtimetermvar}{\trace}$.
With this, property satisfaction is defined as follows:
\bul{whole programs $\wholeprogvar$ satisfy a property $\pi$} iff \iul{$\wholeprogvar$ yields a trace $\trace$} such that \oul{$\trace$ satisifies $\pi$} (\Cref{def:propsat}).

@@

  \begin{nscenter}
    \bul{$\sat{\progvar}{\pi}$}
    $\isdef$
    if \iul{$\forall\runtimetermvar\ \trace,\progstepto{\progvar}{\runtimetermvar}{\trace}$},
    then \oul{$\trace\in\pi$}.
  \end{nscenter}
\end{definition}

@@

This paper uses a more general notion of robust preservation~\cite{abate2021extacc} that applies to compilers that use languages with different trace models. 
To this end, considering a source trace $\src{\trace}$ and a target trace $\trg{\trace}$, there is a relation ($\sim$) describing the effect of a corresponding compiler (see \Cref{sec:casestud:rtp}). 
Using this relation, there are two projection functions: (1) the existential image $\tau_\sim\left(\src{\pi}\right)$ and (2) the universal image $\sigma_\sim\left(\trg{\pi}\right)$.
These projections map source-level (resp. target-level) properties to target-level (resp. source-level) properties. 
The case study of this paper exclusively uses the universal image, since some considered properties, such as \gls*{ss}, are not definable in a higher-level language that, e.g., does not model speculation.
For sake of self-containment, the definition of the universal image is reiterated~\cite{abate2021extacc}:

@@

Using these projection functions, the more general version of robust preservation is defined as follows~\cite{abate2021extacc}:
For a \bul{compiler $\cc{\src{L}}{\trg{L}}$ to robustly preserve a class of source properties $\src{\class}$}, if for any \rul{property $\src{\pi}$ of that class $\src{\class}$ and programs $\src{p}$ written in $\src{L}$} where \iul{$\src{\progvar}$ robustly satisfies $\src{\pi}$}, then \oul{the compilation of $\src{\progvar}$, $\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, must robustly satisfy $\tau_\sim\left(\src{\pi}\right)$}.

\begin{definition}[Robust Preservation with $\tau_\sim$]\label{def:rtp:tau}
  % $\;$\\
  % \vspace{-1em}
  % \begin{nscenter}
  % Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, 
  \bul{$\rtptau{\cc{\src{L}}{\trg{L}}}{\src{\class}}{\sim}$}
  %, iff 
  $\isdef$
  \rul{$\forall \src{\pi}\in\src{\class}, \src{p}\in\src{L},$} if \iul{$\rsat{\src{\progvar}}{\src{\pi}}$}, then \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\tau_\sim{\left(\src{\pi}\right)}}$}.
  % \end{nscenter}
\end{definition}
% 
Note that a class of properties $\class$ can represent just one property $\pi$ by lifting~\cite{clarkson2008hyper} that property to sets of properties, i.e., use the powerset of $\pi$ instead of $\pi$ itself.
Because of this, this paper may write $\rtptau{\cc{\src{L}}{\trg{L}}}{\src{\pi}}{\sim}$, even though $\src{\pi}$ is a property and not a class.
A similar construction can be applied to the projection functions.

Dually to \Cref{def:rtp:tau}, a \bul{compiler $\cc{\src{L}}{\trg{L}}$ to robustly preserves a class of target properties $\trg{\class}$}, if for any \rul{property $\trg{\pi}$ of class $\trg{\class}$ and programs $\src{p}$}, where \iul{$\src{\progvar}$ robustly satisfies $\sigma_\sim\left(\trg{\pi}\right)$}, \oul{the compilation of $\src{\progvar}$, $\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, must robustly satisfy $\trg{\pi}$}.

\begin{definition}[Robust Preservation with $\sigma_\sim$]\label{def:rtp:sigma}
  % $\;$\\
  % \vspace{-1em}
  % \begin{nscenter}
  % Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, 
  \bul{$\rtpsig{\cc{\src{L}}{\trg{L}}}{\trg{\class}}{\sim}$}
  %, iff 
  $\isdef$
  \rul{$\forall \trg{\pi}\in\trg{\class}, \src{p}\in\src{L},$} if \iul{$\rsat{\src{\progvar}}{\sigma_\sim\left(\trg{\pi}\right)}$}, then \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\trg{\pi}}$}.
  % \end{nscenter}
\end{definition}

@@

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g. as in the work of~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls*{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls*{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls*{cf} and \gls*{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls*{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls*{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls*{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls*{cf} and \gls*{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls*{cf} or \gls*{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
Hence, it is of interest to analyze the security of compilation passes and their interation within a compilation pipeline.

@@

  \begin{nscenter}
  \noindent
  \text{\bul{$\wfcsig{\sim}{\trg{\class}}$}} \isdef \text{\rul{$\forall \trg{\Pi}\in\trg{\class}, \sigma_\sim(\trg{\Pi})\in\sigma_\sim(\trg{\class})$}}
  \end{nscenter}
  % $$
\end{definition}

@@

Since the composition of secure compilers is again a secure compiler, the theorems generalise to a whole chain of $n$ secure compilers. 
Moreover, the previously defined theory dually holds for the existential image $\tau_\sim\left(\src{\pi}\right)$, but these results have been moved to the technical report in the interest of space and interest.

\subsection{Secure Compiler Composition with Same Trace Models}
When the cross-language trace relation is an equality, \Cref{thm:rtpsim:sig} collapse:
Given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

\Cref{corr:rtp} provides an easy to use theorem to compose secure compilers without any caveats. 
However, while \Cref{thm:rtpsim:sig} explicitly requires that $\sim_1$ is well-formed w.r.t. $\trg{\class[_2]}$, if $\sim_2$ is not well-formed w.r.t. $\trg{\class[_1]}$, care must be taken. 
%Worst-case is that the resulting class $\sigma_{\sim_1\bullet\sim_2}\left(\trg{\class[_1]}\cap\trg{\class[_2]}\right)$ becomes empty, even though $\trg{\class[_1]}\cap\trg{\class[_2]}$ is not necessarily empty.
\Cref{subsec:compatsecpasses} provides an example of this phenomenon.

%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid w.r.t. security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.
% 
% \MPin{
% 	later on, say that optimisation passes remove elements from traces, so they should be fine w.r.t. the result with $\tau$.
% 	also, yes, if $\tau$ adds we have a problem, if it removes we don't
% \MKin{ Optimisations that remove elements from traces are not necessarily sound,
%  e.g., remove all deallocations to "optimise" memory management by offloading
%  the work to the OS.
% }
% }

@@

@@

This section introduces trace properties of interest for this paper: \gls*{tms}, \gls*{sms}, \gls*{ms}, \gls*{scct}, and \gls*{ss}.
These properties are of practical importance (as mentioned in \Cref{sec:introduction}) and also of interest in the case study (\Cref{sec:casestud:defs,sec:casestud:rtp}) presented later. 
While up to \gls*{scct}, none of the properties is novel, for self-containment of the paper, their definition is necessary for the case study.
For each of the presented properties, the technical appendix accompanying the paper defines corresponding monitors, which refine a given property and have a major role in the proofs of this paper.

@@

For cryptographic code, there is a general guideline that secrets must not be visible on a trace, i.e., marked as $\unlock$.
Moreover, an instruction whose timing is data-dependent must not have a secret as an operand.
Typical operations with data-dependent timing are branches and certain binary operations, such as division. 
\footnote{
	This is architecture-dependent, but division is an operation that serves as a classic example for a data-dependent timing instruction~\cite[p.~755]{arm-refman}.
}
Both operations are also modelled in the trace model by extending the set of base-events with branches on some value $n$ ($\ev{Branch\ n}$) and binary operations with result $n$ ($\ev{Binop\ n}$).

@@

\gls*{scct} may appear overly strict, since it seems that secrets must not occur on a trace (since $\securitytag$ is forced to be $\unlock$). 
However, this is considered standard practice in terms of coding guidelines~\cite{ctguidelines}.
Moreover, programs that operate on secrets and have a ``data independent timing mode'' enabled do not leak these secrets, despite performing computation on them (see \Cref{ex:lscct}).

@@

Essentially, the only work $\sim_{\ctsafe}$ has to do, is to ignore both the new $\ev{Branch\ n}$ and $\ev{Binop\ n}$ base-events as well as the security tag, since they are irrelevant to \gls*{ms}.
The relation is extended from just events to traces in a standard way, skipping the empty event $\emptyevent$ on either side, and it is now possible to define \gls*{msscct} using the universal image:

@@

So far, the considered trace models do not take speculative execution into account. 
For this, extend the earlier trace model (see \Cref{subsec:scct:tracemodel}) so that the security tags ($\securitytag$) carry additional information about the kind of private data leakage, i.e., the type of speculative leak.
Moreover, add pre-events signaling the beginning of a speculative execution ($\ev{Spec}$), a barrier ($\ev{Barrier}$) that signals that any speculative execution may not go past it, as well as a rollback event ($\ev{Rlb}$), which signals that execution resumes to where speculation started.

@@

\gls*{ss}~\cite{patrignani2021exorcising}, similar to \gls*{scct}, is an overapproximation to a variant of noninterference.
The definition of the property itself turns out to be syntactically equal to \gls*{scct}, modulo the underlying trace model.

\begin{definition}[\glsfirst*{ss}]\label{def:trace:ss}
  \noindent\[
    \sssafe:=\left\{\trace[_{\mathghost}] \left|\begin{array}{rcl}
        \trace[_{\mathghost}]=\hole{\cdot} & \text{ or } & \trace[_{\mathghost}]=\preevent[_{\mathghost}];\unlock\cdot\trace[_{\mathghost}'] \wedge \trace[_{\mathghost}']\in\sssafe
                                 \end{array}\right.\right\}
  \]
\end{definition}

@@

The intuition is that speculation is ignored, which amounts to dropping events $\ev{Spec}$, $\ev{Rlb}$, or $\ev{Barrier}$, as well as all base events tagged with $\lock_{\text{PHT}}$. 
Now, define \gls*{specms} as follows:
\begin{definition}[\glsfirst*{specms}]\label{def:trace:specmsdef}
  $
  \specmssafe=\sigma_{\sim_{\mathghost}}\left(\msscctsafe\right)\cap\sssafe
  $
\end{definition}

@@

\vspace{-1.5em}
{\small
\[
  \begin{array}{rrcl}
    \text{(Pre-Events)} & \src{\preevent} &:=& \src{\ev{Alloc \loc\ n}} \mid \src{\ev{Dealloc \loc}} \mid \src{\ev{Get \loc\ n}} \mid \src{\ev{Set \loc\ n}} \\
    \text{(Sandboxtag)} & \src{\sandboxtag} &:=& \src{\ctx} \mid \src{\comp} \\
    \text{(Events)} & \src{\event} &:=& \src{\preevent;\sandboxtag} \mid \src{\emptyevent} \mid \src{\lightning} \\ 
    \text{(Expressions)} & \src{\expr} &:=& \src{x} \mid \src{n} \mid \src{\lbinop{\expr[_1]}{\expr[_2]}} \mid \src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}} \\ 
                         &&&\mid \src{\llet{x}{\expr[_1]}{\expr[_2]}} \mid \src{\llet{x}{new\ \expr[_1]}{\expr[_2]}} \\
                         &&&\mid \src{\ldelete{x}} \mid \src{\lget{x}{\expr}} \mid \src{\lset{x}{\expr[_1]}{\expr[]}} \mid \src{stuck}\\
                         &&&\mid \src{\lcall{f}{\expr}} \mid \src{\lreturn{\expr}} \\
    \text{(Functions)} & \src{F} &:=& (\src{x};\src{\natt\to\natt};\src{\expr}) \\
    \text{(Libraries)} &&& \src{\library_{\ctx}} : \src{Vars} \to \src{F} \hspace{2em}
    \src{\library_{\comp}} : \src{Vars} \to \src{F} \\
  \end{array}
\]
}

>>>>

Properties $\pi$ are sets of traces of admissible program behaviours, ascribing what said property considers valid.
The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}.
A class is simply a set of properties and for the class of safety properties, it is decidable whether a trace satisfies a safety property with just a finite trace prefix.
The compositionality framework (\Cref{sec:sequential}) presented in this paper considers arbitrary classes, while the case-study (\Cref{sec:casestud:rtp}) fixes them to concrete instances.
As an example, consider a trace describing an interaction with a memory where the deallocation of an address $\loc$ precedes a read at that address in memory: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$.
This program behaviour is insecure w.r.t. a canonical notion of (temporal) memory safety dictating no use-after-frees of pointers~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The previous finite trace prefix is enough to decide that the trace does not satisfy \gls*{tms} and there is no way to append events to this prefix which would result in the trace being admissible.

In the following, the execution of a whole program $\wholeprogvar$ that terminates in state $\runtimetermvar$ according to the language semantics and produces trace $\trace$ is written as $\progstepto{\wholeprogvar}{\runtimetermvar}{\trace}$.
With this, property satisfaction is defined as follows:
\bul{whole programs $\wholeprogvar$ satisfy a property $\pi$} iff \iul{$\wholeprogvar$ yields a trace $\trace$} such that \oul{$\trace$ satisifies $\pi$} (\Cref{def:propsat}).

@@

  \begin{nscenter}
    \bul{$\sat{\wholeprogvar}{\pi}$}
    $\isdef$
    if \iul{$\forall\runtimetermvar\ \trace,\progstepto{\wholeprogvar}{\runtimetermvar}{\trace}$},
    then \oul{$\trace\in\pi$}.
  \end{nscenter}
\end{definition}

@@

This paper uses a more general notion of robust preservation~\cite{abate2021extacc} that applies to compilers that use languages with different trace models. 
To this end, considering a source trace $\src{\trace}$ and a target trace $\trg{\trace}$, there is a relation ($\sim$) describing the effect of a corresponding compiler (see \Cref{sec:casestud:rtp}). 
Using this relation, there are two projection functions: (1) the existential image $\tau_\sim\left(\src{\pi}\right)$ and (2) the universal image $\sigma_\sim\left(\trg{\pi}\right)$.
These projections map source-level (resp. target-level) properties to target-level (resp. source-level) properties. 
The case study of this paper exclusively uses the universal image, since some considered properties, such as \gls*{ss}, are not definable in a higher-level language that, e.g., does not model speculation.
For self-containment, the universal image is defined as follows~\cite{abate2021extacc}:

@@

Using these projection functions, the more general version of robust preservation is defined as follows~\cite{abate2021extacc}:
% For a \bul{compiler $\cc{\src{L}}{\trg{L}}$ to robustly preserve a class of source properties $\src{\class}$}, if for any \rul{property $\src{\pi}$ of that class $\src{\class}$ and programs $\src{p}$ written in $\src{L}$} where \iul{$\src{\progvar}$ robustly satisfies $\src{\pi}$}, then \oul{the compilation of $\src{\progvar}$, $\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, must robustly satisfy $\tau_\sim\left(\src{\pi}\right)$}.

% \begin{definition}[Robust Preservation with $\tau_\sim$]\label{def:rtp:tau}
%   % $\;$\\
%   % \vspace{-1em}
%   % \begin{nscenter}
%   % Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, 
%   \bul{$\rtptau{\cc{\src{L}}{\trg{L}}}{\src{\class}}{\sim}$}
%   %, iff 
%   $\isdef$
%   \rul{$\forall \src{\pi}\in\src{\class}, \src{p}\in\src{L},$} if \iul{$\rsat{\src{\progvar}}{\src{\pi}}$}, then \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\tau_\sim{\left(\src{\pi}\right)}}$}.
%   % \end{nscenter}
% \end{definition}
% 
A \bul{compiler $\cc{\src{L}}{\trg{L}}$ to robustly preserves a class of target properties $\trg{\class}$}, if for any \rul{property $\trg{\pi}$ of class $\trg{\class}$ and programs $\src{p}$}, where \iul{$\src{\progvar}$ robustly satisfies $\sigma_\sim\left(\trg{\pi}\right)$}, \oul{the compilation of $\src{\progvar}$, $\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, must robustly satisfy $\trg{\pi}$}.

\begin{definition}[Robust Preservation with $\sigma_\sim$]\label{def:rtp:sigma}
  % $\;$\\
  % \vspace{-1em}
  % \begin{nscenter}
  % Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, 
  \bul{$\rtpsig{\cc{\src{L}}{\trg{L}}}{\trg{\class}}{\sim}$}
  %, iff 
  $\isdef$
  \rul{$\forall \trg{\pi}\in\trg{\class}, \src{p}\in\src{L},$} if \iul{$\rsat{\src{\progvar}}{\sigma_\sim\left(\trg{\pi}\right)}$}, then \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\trg{\pi}}$}.
  % \end{nscenter}
\end{definition}
Note that a class of properties $\class$ can represent just one property $\pi$ by lifting~\cite{clarkson2008hyper} that property to sets of properties, i.e., use the powerset of $\pi$ instead of $\pi$ itself.
Because of this, this paper may write $\rtptau{\cc{\src{L}}{\trg{L}}}{\trg{\pi}}{\sim}$, even though $\trg{\pi}$ is a property and not a class.
A similar construction can be applied to the projection functions.

@@

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g. as in the work of~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls*{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls*{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls*{cf} and \gls*{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls*{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls*{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls*{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls*{cf} and \gls*{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls*{cf} or \gls*{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
Hence, it is of interest to analyze the security of compilation passes and their interaction within a compilation pipeline.

@@

  \begin{nscenter}
  \noindent
  \text{\bul{$\wfcsig{\sim}{\trg{\class}}$}} \isdef \text{\rul{$\forall \trg{\pi}\in\trg{\class}, \sigma_\sim(\trg{\pi})\in\sigma_\sim(\trg{\class})$}}
  \end{nscenter}
  % $$
\end{definition}

@@

Since the composition of secure compilers is again a secure compiler, the theorems generalise to a whole chain of $n$ secure compilers. 
Moreover, the previously defined theory dually holds for the existential image $\tau_\sim\left(\src{\pi}\right)$, but these results have been moved to the technical report in the interest of space.

\subsection{Secure Compiler Composition with Same Trace Models}
When the cross-language trace relation is an equality, \Cref{thm:rtpsim:sig} collapses:
Given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

\Cref{corr:rtp} provides an easy to use way to compose secure compilers without any caveats. 
However, while \Cref{thm:rtpsim:sig} explicitly requires that $\sim_1$ is well-formed w.r.t. $\trg{\class[_2]}$, if $\sim_2$ is not well-formed w.r.t. $\trg{\class[_1]}$, care must be taken. 
%Worst-case is that the resulting class $\sigma_{\sim_1\bullet\sim_2}\left(\trg{\class[_1]}\cap\trg{\class[_2]}\right)$ becomes empty, even though $\trg{\class[_1]}\cap\trg{\class[_2]}$ is not necessarily empty.
\Cref{subsec:compatsecpasses} provides an example of this phenomenon.

%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid w.r.t. security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition, the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.
% 
% \MPin{
% 	later on, say that optimisation passes remove elements from traces, so they should be fine w.r.t. the result with $\tau$.
% 	also, yes, if $\tau$ adds we have a problem, if it removes we don't
% \MKin{ Optimisations that remove elements from traces are not necessarily sound,
%  e.g., remove all deallocations to "optimise" memory management by offloading
%  the work to the OS.
% }
% }

@@

Future work intends to use upper and lower composition as means to build a secure compiler incrementally.

@@

This section introduces trace properties of interest for this paper: \gls*{tms}, \gls*{sms}, \gls*{ms}, \gls*{scct}, and \gls*{ss}.
These properties are of practical importance (as mentioned in \Cref{sec:introduction}) and also of interest in the case study (\Cref{sec:casestud:defs,sec:casestud:rtp}) presented later. 
While up to \gls*{scct}, none of the properties is novel, for self-containment of the paper, their definition is necessary for the case study.
For each of the presented properties, the technical appendix accompanying the paper defines corresponding monitors, which refine a given property and have a key role in the proofs of this paper.

@@

For cryptographic code, there is a general guideline that secrets must not be visible on a trace~\cite{ctguidelines}, i.e., marked as $\unlock$.
In turn, an instruction whose timing is data-dependent must not have a secret as an operand.
Typical operations with data-dependent timing are branches and certain binary operations, such as division. 
\footnote{
	This is architecture-dependent, but division is an operation that serves as a classic example for a data-dependent timing instruction~\cite[p.~755]{arm-refman}.
}
Both operations are also modelled in the trace model by extending the set of base-events with branches ($\ev{Branch\ n}$) and binary operations ($\ev{Binop\ n}$).

@@

\gls*{scct} may appear overly strict, since it seems that secrets must not occur on a trace (since $\securitytag$ is forced to be $\unlock$). 
However, this is considered standard practice in terms of coding guidelines~\cite{ctguidelines}.
Moreover, programs that operate on secrets and have a ``data independent timing mode'' enabled do not leak these secrets, despite performing computation on them~\cite{arm-refman,intel-refman} (see \Cref{ex:lscct}).

@@

Essentially, the only work $\sim_{\ctsafe}$ has to do, is to ignore both the new $\ev{Branch\ n}$ and $\ev{Binop\ n}$ base-events as well as the security tag, since they are irrelevant to \gls*{ms}.
Put differently, traces of \gls*{ms} trivially satisfy \gls*{scct}.

The above relation is extended to traces in a standard way, skipping the empty event $\emptyevent$ on either side, and it is now possible to define \gls*{msscct} using the universal image:

@@

So far, the considered trace models do not take speculative execution into account. 
For this, extend the earlier trace model (see \Cref{subsec:scct:tracemodel}) so that the security tags ($\securitytag$) carry additional information about the kind of private data leakage, i.e., the type of speculative leak.
Moreover, add base-events signaling the beginning of a speculative execution ($\ev{Spec}$), a barrier ($\ev{Barrier}$) that signals that any speculative execution may not go past it, as well as a rollback event ($\ev{Rlb}$), which signals that execution resumes to where speculation started.

@@

\gls*{ss}~\cite{patrignani2021exorcising}, similar to \gls*{scct}, is an overapproximation of a variant of noninterference.
The definition of the property itself turns out to be syntactically equal to \gls*{scct}, modulo the underlying trace model.

\begin{definition}[\glsfirst*{ss}]\label{def:trace:ss}
  \noindent

  \begin{nscenter}
  $
    \sssafe:=\left\{\trace[_{\mathghost}] \left|\begin{array}{rcl}
        \trace[_{\mathghost}]=\hole{\cdot} & \text{ or } & \trace[_{\mathghost}]=\preevent[_{\mathghost}];\unlock\cdot\trace[_{\mathghost}'] \wedge \trace[_{\mathghost}']\in\sssafe
                                 \end{array}\right.\right\}
  $ 
  \end{nscenter}
\end{definition}

@@

The intuition is that \gls*{ss} is trivially satisfied in \gls*{scct}, since speculation is inexpressible there, which amounts to dropping events $\ev{Spec}$, $\ev{Rlb}$, or $\ev{Barrier}$, as well as all base events tagged with $\lock_{\text{PHT}}$.

Now, define \gls*{specms} as: 
\begin{definition}[\glsfirst*{specms}]\label{def:trace:specmsdef}
  $
  \specmssafe=\sigma_{\sim_{\mathghost}}\left(\msscctsafe\right)\cap\sssafe
  $
\end{definition}

@@

\vspace{-1.5em}
{\small
\[
  \begin{array}{rrcl}
    \text{(Base-Events)} & \src{\preevent} &:=& \src{\ev{Alloc \loc\ n}} \mid \src{\ev{Dealloc \loc}} \mid \src{\ev{Get \loc\ n}} \mid \src{\ev{Set \loc\ n}} \\
    \text{(Sandboxtag)} & \src{\sandboxtag} &:=& \src{\ctx} \mid \src{\comp} \\
    \text{(Events)} & \src{\event} &:=& \src{\preevent;\sandboxtag} \mid \src{\emptyevent} \mid \src{\lightning} \\ 
    \text{(Expressions)} & \src{\expr} &:=& \src{x} \mid \src{n} \mid \src{\lbinop{\expr[_1]}{\expr[_2]}} \mid \src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}} \\ 
                         &&&\mid \src{\llet{x}{\expr[_1]}{\expr[_2]}} \mid \src{\llet{x}{new\ \expr[_1]}{\expr[_2]}} \\
                         &&&\mid \src{\ldelete{x}} \mid \src{\lget{x}{\expr}} \mid \src{\lset{x}{\expr[_1]}{\expr[]}} \mid \src{stuck}\\
                         &&&\mid \src{\lcall{f}{\expr}} \mid \src{\lreturn{\expr}} \\
    \text{(Functions)} & \src{F} &:=& (\src{x};\src{\natt\to\natt};\src{\expr}) \\
    \text{(Libraries)} &&& \src{\library_{\ctx}} : \src{Vars} \to \src{F} \hspace{2em}
    \src{\library_{\comp}} : \src{Vars} \to \src{F} \\
  \end{array}
\]
}
